Expanding tidy data principles to facilitate missing data exploration, visualization and assessment of imputations

### 2020-01-09 paper/tidy-missing-data-paper.Rmd
tidy up section 3

@@

Imputation methods are more accessible than they have ever been, and the number of methods and implementations in software continues to grow. Values can be imputed with one value (single imputation), or multiple values (multiple imputation), creating $m$ datasets. Methods and software for single and multiple imputation are now discussed.

### Single imputation {#single-imputation}

\pkg{VIM} [@VIM] efficiently implements well-used imputation methods K nearest neighbors, regression, hot-deck, and iterative robust model-based imputation. This diversity of approaches allows the package to deal with semi-continuous, continuous, count, and categorical data. VIM identifies imputed cases by adding an indicator variable with a suffix `_imp`. So `Var1` would have a sibling indicator column, `Var1_imp`, with TRUE or FALSE values to indicate imputed value. \pkg{VIM} also has a variety of visualization methods, discussed in section \@ref(exploration). The \pkg{simputation} package provides an interface into popular imputation methods for linear regression, decision trees, K nearest neighbors, hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977]. Other software for imputation includes \pkg{Hmisc} [@Hmisc], which provide predictive mean matching, \pkg{imputeTS} [@imputeTS], which provide time series imputation methods, and \pkg{missMDA} [@missMDA], which imputes data using principal components analysis.

A key feature of the design of \pkg{simputation} is that it returns a regular dataframe. This is different from other imputation packages, which return custom classes of data or additional objects. These create extra work when preparing for subsequent analysis. Providing a dataframe with imputed values included in the data reduces the friction of working with other tools, but comes at the cost of identifying imputed values.

### Multiple imputation {#multiple-imputation}

Multiple imputation is often regarded as best practice for imputing values [@Schafer2002], as long as appropriate caution is taken [@Sterne2009]. Popular and robust methods for multiple imputation include the \pkg{mice}, \pkg{Amelia}, and \pkg{mi} packages [@mice; @amelia; @mi]. The \pkg{mice} package implements the method of chained equations, using a variable-wise algorithm to calculate the posterior distribution of parameters to generate imputed values. The workflow in \pkg{mice} revolves around imputing data, returning completed data, and fitting a model and pooling the results.

The \pkg{Amelia} package [@amelia] assumes data are multivariate normal, and samples from the posterior using the computationally efficient (and parallelizable) Expectation-Maximization Bootstrap (EMB) algorithm [@Honaker2010] and allows for incorporation of information on the values in a prior.

The \pkg{mi} package [@mi] also uses Bayesian models for imputation, providing better handling of semi-continuous values, and data with structural or perfect correlation. A collection of analysis models are also provided in \pkg{mi}, which work with the specially multiply imputed data, including linear models, generalized linear models, and their corresponding Bayesian components. This approach promotes fluid workflow, with a similar penalty to tidying up model output, which is still messy.

Another software for performing multiple imputation is \pkg{norm} [@norm; @schafer-norm], which uses methods from the NORM software from @schafer-norm, such as multiple imputations using EM for multivariate normal data. The \pkg{norm} package does not provide a framework for tracking missing values, but instead provides tools for making inference from multiple imputation.

Each of these multiple imputation methods provide practical methods, but all have different output structures, and do not have consistent interfaces in their implementation. This makes then inherently messy and challenging to integrate into an analysis pipeline. An ideal approach would have consistent data structures that work with other analysis tools, and track missing values by default. This would make the data outputs tidy, streamlining subsequent analysis.

### Exploration {#exploration}

The primary focus of most missing data packages is making inferences, and exploring imputed values, not
on exploring relationships in missing values, and identifying possible patterns.
Texts covering exploration of missing data have the same problem as with modelling: the input is tidy, but the output does not work with other tools [@VanBuuren2012]; this is inefficient. Methods for exploring missing values are primarily covered in literature on interactive graphics [@Swayne1998; @Unwin1996; @Cook2007], and are picked up again in a discussion of a graphical user interface [@Cheng2015].

The missingness matrix $R$ can be used to assess missing data dependence, and has  been used in interactive graphics, dubbed a "shadow matrix", to link missing and imputed values to the data, facilitating their display in graphics [@Swayne1998]. This work also focuses heavily on multivariate numeric data. This is an idea upon which this new work builds.

The approach of the MANET software in @Unwin1996 focussed on multivariate categorical data, where missingness is explicitly added as an additional category. MANET also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot. The MANET software is no longer maintained and cannot be installed. The approach of @Swayne1998 in the software XGobi, further developed in ggobi [@Cook2007], focussed more on multivariate quantitative data. Missingness is incorporated into plots in ggobi by setting them to be 10% below the minimum value.

MissingDataGUI provides a user interface for exploring missing data structure both numerically and visually. Using a Graphical User Interface (GUI) to explore missing data may make it easier to glean valuable insight into important structures for missingness. However, this comes as a trade off with GUIs - the insights are typically not captured or recorded, so it is difficult to incorporate them into reproducible analyses. This distracts and breaks analysis workflow, inviting mistakes.

VIM (Visualizing and Imputing Missing Data) provides visualization methods to identify observed, imputed, and missing values. These include spinograms, spinoplots, missingness matrices, plotting missingness in the margins of other plots, and other summaries. These help explore missingness, however these visualizations do not map variables to graphical aesthetics, creating friction moving across workflows, making them difficult to extend to new circumstances. Additionally, data used to create the visualizations cannot be accessed, posing a barrier to further explorations.

\pkg{ggplot2} incorporates missingness into visualizations only when mapping a discrete variable to a graph aesthetic. This has some limitations. For example, for data with school grades and test scores, a boxplot visualization could have school grade mapped to the x axis, and test score to the y axis (See Figure \@ref(fig:gg-box-na)). If there are missings in a continuous variable like test score, \pkg{ggplot2} omits the missing values and prints a warning message. However, if a discrete variable like school year has missing values, an NA category is created for school year, and the scores are placed into this (Figure \@ref(fig:gg-box-na)).

@@

```{r gg-box-na, fig.height = 3, fig.width = 9, out.width = "100%", fig.align = "center", fig.cap = "ggplot2 provides different visualizations depending on what type of data has missing values. (A) Data is complete and graphic is presented; (B) Missings are only in category variable year - an additional 'NA' boxplot is created; (C) Missings only in scores, no additional missingness information is shown; (D) Missings in both scores and school year, additional missing information is shown. The missingness category is only shown when there are missings in categorical variables such as year (plots (B) and (D)). In (C), no missingness information is given on the graphic, despite there being missings in score, and a warning message is displayed about the number of missing values omitted.", fig.show = "asis"}

>>>>

Methods for exploring, understanding, and imputing missing data are more accessible than they have ever been. In particular, the number of methods and implementations in imputation seeing continued growth. Values can be imputed with one value (single imputation), or multiple values (multiple imputation), creating $m$ datasets. This section discusses existing software for single and multiple imputation \@ref(imputation), and exploration \@ref(exploration).

@@

\pkg{VIM} [@VIM] implements well-used imputation methods K nearest neighbors, regression, hot-deck, and iterative robust model-based imputation. These diverse approaches allows for imputing with semi-continuous, continuous, count, and categorical data. VIM identifies imputed cases by adding an indicator variable with a suffix `_imp`. So `Var1` has a sibling column, `Var1_imp`, with values TRUE or FALSE indicating imputation. \pkg{VIM} also has a variety of visualization methods, discussed in \@ref(exploration). \pkg{simputation} provides an interface to imputation methods from VIM, in addition providing hotdeck imputation, and the EM algorithm [@simputation; @Dempster1977]. \pkg{simputation} provides a consistent formula interface to all imputation, and always returns a dataframe with the updated imputations. \pkg{Hmisc} [@Hmisc], provide predictive mean matching, \pkg{imputeTS} [@imputeTS], provides time series imputation methods, and \pkg{missMDA} [@missMDA], imputes data using principal components analysis.

Multiple imputation is often regarded as best practice for imputing values [@Schafer2002], as long as appropriate caution is taken [@Sterne2009]. Popular and robust methods for multiple imputation include the \pkg{mice}, \pkg{Amelia}, and \pkg{mi} packages [@mice; @amelia; @mi]. \pkg{mice} implements the method of chained equations, using a variable-wise algorithm to calculate the posterior distribution of parameters to generate imputed values. The workflow in \pkg{mice} revolves around imputing data, returning completed data, and fitting a model and pooling the results.

\pkg{Amelia} [@amelia] assumes data are multivariate normal, and samples from the posterior using the computationally efficient (and parallelizable) Expectation-Maximization Bootstrap (EMB) algorithm [@Honaker2010] and allows for incorporation of information on the values in a prior.

\pkg{mi} [@mi] also uses Bayesian models for imputation, providing better handling of semi-continuous values, and data with structural or perfect correlation. A collection of analysis models are also provided in \pkg{mi}, which work with the specially multiply imputed data, including linear models, generalized linear models, and their corresponding Bayesian components. This approach promotes fluid workflow, with a similar penalty to tidying up model output, which is still messy.

\pkg{norm} [@norm], provides multiple imputation using EM for multivariate normal data, drawing from methods in the NORM software [@schafer-norm]. \pkg{norm} does not provide a framework for tracking missing values, instead providing tools for making inference from multiple imputation.

### Summary {#imputation-summary}

Each imputation method provides practical methods for different usecases, but most have different output structures, and do not have consistent interfaces in their implementation. This makes then inherently messy and challenging to integrate into an analysis pipeline. For example, combining different imputation methods from different pieces of software is not currently straightforward. \pkg{simputation} is different from other imputation packages, and resolves some of these complications with a simple approach of a unified syntax for all imputation, and always returning a dataframe containing imputed values. This reduces the friction of working with other tools, but comes at the cost of identifying imputed values. An ideal approach would use consistent, simple data structures that work with other analysis tools, and help track missing values. This would make imputation outputs tidy, streamlining subsequent analysis.

## Exploration {#exploration}

The primary focus of most missing data packages is making inferences, and exploring imputed values, not
on exploring relationships in missing values, and identifying possible patterns.
Texts covering the exploration phase of missing data have the same problem as with modelling: the input is tidy, but the output does not work with other tools [@VanBuuren2012]; this is inefficient. Methods for exploring missing values are primarily covered in literature on interactive graphics [@Swayne1998; @Unwin1996; @Cook2007], and are picked up again in a discussion of a graphical user interface [@Cheng2015].

The missingness matrix $R$ can be used to assess missing data dependence It has been used in interactive graphics, dubbed a "shadow matrix", to link missing and imputed values to the data, facilitating their display [@Swayne1998], focusing heavily on multivariate numeric data. This is an idea upon which this new work builds.

The MANET (Missings Are Now Equally Treated) software @Unwin1996 focussed on multivariate categorical data, with missingness explicitly added as a category. MANET also provided univariate visualizations of missing data using linked brushing between a reference plot of the missingness for each variable, and a plot of the data as a histogram or barplot. The MANET software is no longer maintained and cannot be installed. The approach of @Swayne1998 in the software XGobi, further developed in ggobi [@Cook2007], focussed more on multivariate quantitative data. Missingness is incorporated into plots in ggobi by setting them to be 10% below the minimum value.

MissingDataGUI provides a Graphhical User Interface (GUI) for exploring missing data structure both numerically and visually. Using a GUI to explore missing data facilitates rapid insight into missingness structures. However, this comes as a trade off, as insights are not captured or recorded with a GUI, making it challenging to incorporate into reproducible analyses. This distracts and breaks analysis workflow, inviting mistakes.

VIM (Visualizing and Imputing Missing Data) provides visualization methods to identify and explore observed, imputed, and missing values. These include spinograms, spinoplots, missingness matrices, plotting missingness in the margins of other plots, and other summaries. However, these visualizations do not map variables to graphical aesthetics, creating friction when moving through analysis workflows, making them difficult to extend to new circumstances. Additionally, data used to create the visualizations cannot be accessed, posing a barrier to further exploration.

\pkg{ggplot2} incorporates missingness into visualizations only when mapping a discrete variable to a graph aesthetic. This has some limitations. For example, data of school grades and test scores, a boxplot visualization with school grade and test score mapped to the x- and y-axis (Figure \@ref(fig:gg-box-na)). If there are missings in a continuous variable like test score, \pkg{ggplot2} omits the missings and prints a warning message. However, if a discrete variable like school year has missing values, an NA category is created for school year, where scores are placed (Figure \@ref(fig:gg-box-na)).

@@

```{r gg-box-na, fig.height = 3, fig.width = 9, out.width = "100%", fig.align = "center", fig.cap = "ggplot2 provides different visualizations depending on what type of data has missing values for data of student test scores in school year. (A) Data is complete; (B) Missings are only in year - an additional 'NA' boxplot is created; (C) Missings only in scores, no additional missingness information is shown; (D) Missings in both scores and year, additional missing information is shown. The missingness category is only shown when there are missings in categorical variables such as year (plots (B) and (D)). In (C), no missingness information is given on the graphic, despite there being missings in score, and a warning message is displayed about the number of missing values omitted.", fig.show = "asis"}
